\section{Introduction}

\subsection{Contributions}

In this paper we describe our experiment in implementing an algebraic effect
for safe concurrent resource management, which we call \code{ConcState}.

The effect is implemented using the Effects library in Idris and is modelled
after the \code{LockLang} eDSL~\cite{cbconc-fi} by Edwin Brady and Kevin
Hammond.

The full code, together with instructions for running it, is available at\\
\mbox{\url{https://github.com/yfyf/idris-concstate-effect}}.\\
It has been tested to work with Idris version \version.

\subsection{Motivation}

Concurrency is a concept which is very hard to simulate in our brains.
Rephrasing my supervisor: if sequential programs have a linear state space of
$O(n)$, then concurrent programs have an exponential state space of $O(2^n)$.
This is perhaps our main struggle with concurrent code.

Historically, concurrent protocols and algorithms were implemented either in an
(unsafe) ad-hoc manner or formal models were employed for proving the
correctness and safety properties. In the latter case, a formal descriptions
is usually ``manually'' translated into an actual implementation. Due to the
subtle nature of concurrent algorithms, the implementation often turns out to
be unfaithful to the formal description, resulting in hard to trace bugs.

Naturally, both approaches are quite unsustainable. It is a shame that, despite
lots of research in related areas, in practice most software systems handle
concurrent resources without any static safety guarantees (e.g. without any
hints from the compiler about deadlocks).

Hence the initial aim of this project was to explore advances in programming
language research in order to address these problems. This idea crystallised
into an attempt to leverage the power of dependent types, in particular, the
way they are used in Idris. Our main goal was to combine previous work on
concurrent resource handling~\cite{cbconc-fi} in Idris with algebraic effects
(as they are implemented in the Effects~\cite{effects-idr} library).

\subsection{Dependent types and type safety}

Usually, programming languages treat types and values as completely distinct
entities. Most of the time, types are somewhat ``static'' and can't be
manipulated much, except for basic constructions, such as products and unions.

Some (mostly functional) languages are more powerful in this respect. For
example, in Haskell one has polymorphic types and type constructors which act a
lot like functions: they can take as input multiple type arguments and can be
partially applied.

Dependent types are on the opposite side of the spectrum, where the distinction
between types and values is blurred and it is possible to construct
\emph{types} which depend on \emph{values}. A more powerful type system allows
the programmer to more accurately describe both data types and program flow.
This implies that static checking performed by the compiler can provide greater
correctness guarantees. In fact, sufficiently precise typing ensures that a
program which type checks is actually correct.

A typical example of a data structure possible with dependent types is a
fixed-length list, i.e. a vector. It can be expressed as a Haskell-style GADT
as follows:

\begin{BVerbatim}

-- the Vect is parametrised over a length and a type
data Vect : Nat -> Type -> Type where
    -- empty vector constructor, Z stands for 0
    Nil  : Vect Z a
    -- the standard cons operator, where `S` is
    -- the natural number successor operator
    (::) : a -> Vect n a -> Vect (S n) a

\end{BVerbatim}

which makes it possible to describe a ``safe'' version of the standard
list-type function \code{head} and give a precise type for appending (\code{++}):

\begin{BVerbatim}

-- ensure the vector is non-empty by matching the length as (S n)
head : Vect (S n) a -> a
head (x::xs) = x

-- carry around the vector length when appending
(++) : Vect m a -> Vect n a -> Vect (m + n) a
(++) []      ys = ys
(++) (x::xs) ys = x :: xs ++ ys

\end{BVerbatim}

We can see that the type of \code{head} ensures it is a \emph{total} function
(on non-empty vectors). This is an example of how dependent types help ensure
our programs are correct: a run-time error of \code{head} being applied to an
empty vector is ruled out at compile time, since type unification would simply
fail in an application of \code{head} on a possibly-empty vector.

For a quick introduction to dependent types and a list of great references we
refer the reader to Jan Malakhovski's introduction~\cite{deptypes-intro}.

\subsection{Idris}

\Idris{} is a research language with full dependent types conceived by Edwin
Brady at the University of St Andrews. Idris is influenced by Haskell and tries
to be a general purpose language aimed at systems programming (via easy
interfacing with C, compilation to multiple backends (C, JavaScript, Java) and
various performance optimisations).

Since our goal was to make writing safe \emph{real-world} concurrent code
easier, Idris was a good match for an experimentation platform.

For an introduction to Idris please consult the Idris
tutorial~\cite{idris-tutorial}. Finer details about the core Idris can be found
in the implementation paper~\cite{idris-impl}.
